---
title: 'Machine learning - Lasso Regression of hot and cold cancers'
author: 'Joshua Zeng'
date: today
bibliography: references.bib
format:
  html:
    code-fold: true
---

# Two classes (e.g. cases vs. controls)

In this exercise, we will analyze a data set from @patel2021.
The data set is available at [this link](https://www.ebi.ac.uk/biostudies/files/S-BSST416/COVID19%20blood%20biomarker%20profile%20with%20phenotype.xlsx).
It can be loaded into R and cleaned by the commands below.


```{r}
suppressPackageStartupMessages({
  library(arrow)
  library(glue)
  library(rio)
  library(dplyr)
  library(glmnet)
  library(ggplot2)
  library(randomForest)
})
```


```{r}
data_dir = '/media/josh/wd/data/ucsc_xena_data'
out_dir = '/media/josh/wd/data/tcga_data'
cancer_type <- 'BRCA'
```

```{r}

rna = data.table::fread(glue('{data_dir}/tcga_RSEM_gene_tpm.gz'))
```

```{r}
rna <- as.data.frame(rna)
row.names(rna) <- rna$sample
rna$sample <- NULL
rna[1:3, 1:7]
```
```{r}
genes = read.csv('/media/josh/wd/data/ensembl_hgnc/ensembl_ncbi_gene_ID.from_bioMart.csv')
genes <- genes %>% filter(gene_biotype == 'protein_coding') %>% filter(!is.na(hgnc_symbol))
genes <- genes[!duplicated(genes$hgnc_symbol), ]
genes = genes[, c('ensembl_gene_id', 'hgnc_symbol')]
dim(genes)
```


```{r}
meta = readxl::read_excel(glue('{out_dir}/TIL_patterns_from_4759_TCGA_subjects.xlsx'))

meta_sub= meta %>% filter(Study == !!cancer_type)
table(meta_sub$Global_Pattern)

```

```{r}
meta_sub = meta_sub %>% filter(Global_Pattern %in% c('Non-Brisk Focal', 'Non-Brisk Multifocal', 'Brisk Diffuse'))
table(meta_sub$Global_Pattern)
```

```{r}
meta_sub$Class = ifelse((meta_sub$Global_Pattern == 'Brisk Diffuse'), 'hot', 'cold') 
table(meta_sub$Class)
```

```{r}
# Subset primary solid tumor
tcga_meta = data.table::fread(glue('{data_dir}/Survival_SupplementalTable_S1_20171025_xena_sp', sep='\t')) %>% as.data.frame()
tcga_meta = tcga_meta %>% filter(`cancer type abbreviation` == !!cancer_type)
tcga_meta$sample_code <-  sapply(strsplit(tcga_meta$sample, '-'), '[', 4)
tcga_meta <- tcga_meta %>% filter(sample_code == '01')
tcga_meta[1:3, 1:5]
```

```{r}
meta_merged = merge(tcga_meta, meta_sub, by.x = '_PATIENT', by.y = 'ParticipantBarcode')
```

```{r}
# Subset the samples to the BRCA dataset
dim(rna)
rna_sub <- rna[, intersect(colnames(rna),  meta_merged$sample)]
dim(rna_sub)
# Remove the suffix in the gene names
row.names(rna_sub) = sapply(strsplit(row.names(rna_sub), '\\.'), '[', 1)
# keep protein-coding genes
# rna_sub = rna_sub.loc[[gene_dict[g] for g in rna_sub.index.tolist() if g in genes['ensembl_gene_id'].tolist()], :]

rna_sub = merge(genes, rna_sub, by.x = 'ensembl_gene_id', by.y = 'row.names')
row.names(rna_sub) = rna_sub$hgnc_symbol
rna_sub = rna_sub %>% select(- c('ensembl_gene_id', 'hgnc_symbol'))
dim(rna_sub)
```

```{r}
rna_sub_t = t(rna_sub)
rna_sub_t[1:3, 1:3]
```

```{r}
meta_df = meta_merged[, c('sample', 'Class')]
head(meta_df)
```

```{r}
clean_df = merge(meta_df, rna_sub_t, by.x = 'sample', by.y = 'row.names')
dim(clean_df)
```

```{r}
row.names(clean_df) = clean_df$sample
clean_df = clean_df %>% select(-c('sample'))
clean_df[1:3, 1:3]
```

```{r}
table(clean_df$Class)
```



Here, we will classify the patients with hot and cold tumors


```{r}
set.seed(88)
n_sample <- nrow(clean_df)  # number of samples
idx_rnd <- sample(n_sample)    # randomly sampled indices
cutoff_0.8 <- floor(n_sample * 0.8)
idx_train <- idx_rnd[1:cutoff_0.8]
idx_test  <- idx_rnd[(cutoff_0.8 + 1):n_sample]

df_train <- clean_df[idx_train, ]
df_test  <- clean_df[idx_test, ]
print(table(df_train$Class))
print(table(df_test$Class))
```

## Lasso logistic regression

Reformat the data frame to a matrix and a response variable using `model.matrix`. 
In the matrix and the response variable, categorical variables are converted to dummy variables having numeric values.

```{r}
# options(expressions = 5e5)
options("exppression" = 500000)
# memory.limit(size=8000000)
# x_train <- model.matrix(Class ~ ., df_train)[, -1]   # no (Intercept)

x_train <- df_train %>% select(-c('Class'))

# x_train <- makeX(x_train, na.impute = T)
x_train <- as.matrix(x_train)

scaled.x_train = scale(x_train)
genes.with.na <- colnames(scaled.x_train)[colSums(is.na(scaled.x_train)) > 0]
scaled.x_train <- scaled.x_train[, !colnames(scaled.x_train) %in% genes.with.na]

# colnames(scaled.x_train)[colSums(is.na(scaled.x_train)) > 0]
# x_train <- model.matrix(Class ~ ., df_train)[, -1] 

y_train <- ifelse(df_train$Class == 'hot', 1, 0)
```

```{r}
print(table(df_train$Class))
print(table(y_train))
```

3. Find the optimal $\lambda$ from cross-validation using `cv.glmnet`.

```{r}
cv.lasso <- cv.glmnet(
  x = scaled.x_train, 
  y = y_train, 
  alpha = 1,               # lasso
  family = 'binomial'      # logistic
)
optimal_lambda <- cv.lasso$lambda.min
plot(cv.lasso)

```

4. Fit a lasso logistic model to the training data with the optimal $\lambda$.
Find which variables got non-zero coefficients using `coef`.

```{r}
lasso_logit <- glmnet(
  x = scaled.x_train, 
  y = y_train, 
  alpha = 1,               # lasso
  family = 'binomial',     # logistic
  lambda = optimal_lambda
)
colnames(scaled.x_train)[predict(lasso_logit, type = "nonzero")[, 1]]

lasso_logit_coef <- coef(lasso_logit, optimal_lambda)
lasso_logit_coef[lasso_logit_coef[, 1] != 0, ]
```

```{r, fig.height= 7, fig.width= 7}
plot_df <- lasso_logit_coef[lasso_logit_coef[, 1] != 0, ] %>% as.data.frame()
plot_df$feature <- row.names(plot_df)
names(plot_df)[1] <-  'coefficient'
plot_df <- plot_df %>% filter(feature != '(Intercept)') %>% arrange(desc(coefficient))
plot_df$feature <- factor(plot_df$feature, levels = unique(plot_df$feature))
head(plot_df)
features.select <- plot_df$feature
ggplot(data=plot_df,                       # plot coefficients
       aes(x=feature, y=coefficient)) +
       coord_flip() +         
       geom_bar(stat='identity',   fill= 'royalblue') +
       guides(fill=FALSE) 
```

```{r}
x_test <- df_test %>% select(-c('Class'))

# x_test <- makeX(x_test, na.impute = T)
x_test <- as.matrix(x_test)
# genes.with.na <- colnames(x_test )[colSums(is.na(x_test)) > 0]
x_test <- x_test[, !colnames(x_test) %in% genes.with.na]

scaled.x_test = scale(x_test)
# colnames(scaled.x_test)[colSums(is.na(scaled.x_test)) > 0]
# x_test <- model.matrix(Class ~ ., df_test)[, -1] 

```


5. Draw the ROC curve for predicting the test set and compute the AUC. 
There are several packages with functions for ROC curves and AUC computation, e.g. `pROC`, `ROCR` and `yardstick`. 
Note, the solution uses `roc_curve` and `roc_auc_vec` in the `yardstick` package in the `tidymodels` framework.

```{r}
# x_test <- model.matrix(Group ~ ., df_test)[, -1]   # no (Intercept)
y_test <- ifelse(df_test$Class == 'hot', 1, 0)
y_pred <- predict(lasso_logit, newx = x_test, type = "response")[, 1]
y_test_df <- data.frame(truth = factor(y_test, levels = c(1, 0)), pred = y_pred)
print(y_test_df)
roc <- yardstick::roc_curve(y_test_df, truth, pred)
print(roc)

ggplot(roc) +
  aes(x = 1 - specificity, y = sensitivity) +
  geom_path() +
  geom_abline(lty = 3) +
  coord_equal()
auc <- yardstick::roc_auc_vec(factor(y_test, levels = c(1, 0)), y_pred)
cat("AUC =", auc)
```

Using `autoplot` for ROC curve,

```{r}
autoplot(roc)
```


## Random forest


1. Fit a random forest model to the training data and find its error rate 

```{r}
rf <- randomForest(x = scaled.x_train[, features.select], y = factor(y_train), importance=TRUE)
print(rf)
```

3. Make the ROC curve for predicting the test set and compute the AUC. 
There are several packages with functions for ROC curves and AUC computation, e.g. `pROC`, `ROCR` and `yardstick`. 
Note, the solution uses `roc_curve` and `roc_auc_vec` in the `yardstick` package in the `tidymodels` framework.

```{r}
# x_test <- df_test[, -1]
# scaled.x_test
# y_pred <- predict(lasso_logit, newx = x_test, type = "response")[, 1]
# y_pred <- predict(rf, newdata = scaled.x_test[, features.select], type = "prob")
y_pred <- predict(rf, newdata = scaled.x_test[, features.select], type = "response")
# y_test_df <- data.frame(truth = df_test[[1]], pred = y_pred)
y_test_df <- data.frame(truth = factor(y_test, levels = c(0, 1)), pred = y_pred)
y_test_df$pred <- factor(y_test_df$pred, levels = c(0, 1))
# y_test_df$truth <- as.numeric(as.character(y_test_df$truth))
# y_test_df$pred <- as.numeric(as.character(y_test_df$pred))

print(y_test_df)
yardstick::conf_mat(y_test_df, truth, pred)
roc <- yardstick::roc_curve(y_test_df, truth, pred)
print(roc)
ggplot(roc) +
  aes(x = 1 - specificity, y = sensitivity) +
  geom_path() +
  geom_abline(lty = 3) +
  coord_equal()

auc <- yardstick::roc_auc_vec(factor(y_test, levels = c(1, 0)), y_pred)
cat("AUC =", auc)

```

Using `autoplot` for ROC curve,

```{r}
autoplot(roc)
```


# Multi-classes (e.g. class 1, 2 and 3)

In this exercise, we will re-analyze a data set from `mixOmics` package.
According to the authors: 
“The Small Round Blue Cell Tumors (SRBCT) dataset from includes the expression levels of 2,308 genes measured on 63 samples. The samples are classified into four classes as follows: 8 Burkitt Lymphoma (BL), 23 Ewing Sarcoma (EWS), 12 neuroblastoma (NB), and 20 rhabdomyosarcoma (RMS).”

The dataset can be downloaded from [this link](https://github.com/mixOmicsTeam/mixOmics/raw/master/data/srbct.rda) and loaded into R by the following commands.

```{r}
#| code-fold: false
data(srbct, package = 'mixOmics')

# When `mixOmics` package was not installed.
tmp_file <- tempfile(fileext = "rda")
download.file("https://github.com/mixOmicsTeam/mixOmics/raw/master/data/srbct.rda", tmp_file)
load(tmp_file)    # loads `srbct`
```

After loading the data, we will have a variable called `srbct` which is a list containing the following:

* `gene`: a data frame with 63 rows and 2308 columns. The expression levels of 2,308 genes in 63 subjects.
* `class`: a class vector containing the class tumor of each individual (4 classes in total).
* `gene.name`: a data frame with 2,308 rows and 2 columns containing further information on the gene

We can combine `gene` and `class` into a single data frame using:

```{r}
#| code-fold: false
srbct_df <- data.frame(class = srbct$class, srbct$gene)
# optional, if you prefer `tibble` in `tidyverse` package set to `data.frame`
library(tidyverse)
srbct_df <- as_tibble(srbct_df)
```

We are going to use random forests to find variables that are important for discriminating the 4 classes.

1. Load the `randomForest` R package.

```{r}
library(randomForest)
```

2. Randomly split your data into a training (80% of the data) and testing sets (20% of the data).

```{r}
n_sample <- nrow(srbct_df)  # number of samples
idx_rnd <- sample(n_sample)    # randomly sampled indices
cutoff_0.8 <- floor(n_sample * 0.8)
idx_train <- idx_rnd[1:cutoff_0.8]
idx_test <- idx_rnd[(cutoff_0.8 + 1):n_sample]

df_train <- srbct_df[idx_train, ]
df_test <- srbct_df[idx_test, ]
```

3. Tune a hyper-parameter (`mtry`) of random forest (only on training data) using `tuneRF` function in `randomForest` package.

```{r}
y_train <- df_train[[1]]
x_train <- df_train[, -1]
rf_tune_res <- tuneRF(x = x_train, y = y_train, mtryStart = 100, ntreeTry = 100)
print(rf_tune_res)
```

4. Fit a random forest model to the training data and find its error rate (either out of bag error or cross validation. up to you!)

```{r}
rf <- randomForest(x = x_train, y = y_train, mtry = 200)
print(rf)
```

5. What is the accuracy of predicting the test set?

```{r}
x_test <- df_test[, -1]
y_pred <- predict(rf, newdata = x_test)
yardstick::accuracy_vec(df_test[[1]], y_pred)
```

6. Find the top 10 most important genes for discriminating all the classes

```{r}
rf <- randomForest(x = x_train, y = y_train, mtry = 200, importance = TRUE)
print(rf)
imps_mda <- randomForest::importance(rf, type = 1)
print(head(imps_mda))
imps_mda <- imps_mda[, 1]
imps_mda[order(imps_mda, decreasing = TRUE)][1:10]
varImpPlot(rf)
```

7. What are the top 10 genes for predicting _EWS_ class

```{r}
imps_ews <- importance(rf, type = 1, class = "EWS")[, 1]
imps_ews[order(imps_ews, decreasing = TRUE)][1:10]
varImpPlot(rf, type = 1, class = "EWS")
```



